Formalization of Typed Lua

### 2015-06-11 doc/papers/paper/paper.tex
rewrite of the beginning of section on typing rules, renaming of typing rules, DLS paper

In this section we use a reduced core of Typed Lua to present
the typing rules that type check the novel features of our type system.
These rules type check table refinement and
overloading on the return type of functions.

Our core limits control flow to if and while statements;
it has explicit type annotations, explicit scope for variables,
explicit method declarations, and explicit method calls.
Here is a list of features that are not present in our reduced core:
\begin{itemize}
\item labels and goto statements (they are difficult to handle along
with our simplified form of \emph{flow typing}, and they are out of
scope for now);
\item explicit blocks (we are already using explicit scope for variables);
\item other loop structures such as repeat-until, numeric for,
and generic for (we can use while to express them);
\item table fields other than $[e_{1}] = e_{2}$
(we can use this form to express the missing forms); 
\item arithmetic operators other than $+$
(other arithmetic operators have similar typing rules);
\item relational operators other than $==$ and $<$
(inequality has similar typing rules to $==$ and
other relational operators have similar typing rules to $<$);
\item bitwise operators other than $\&$
(other bitwise operators have similar typing rules).
\end{itemize}

Our reduced core does not lose much expressiveness, as it can express
any Lua program except those that use labels and goto statements.

\begin{figure*}[!ht]
\textbf{Abstract Syntax}\\
\dstart
$$
\begin{array}{rlr}
s ::= & & \textsc{statements:}\\
& \;\; \mathbf{skip} & \textit{skip}\\
& | \; s_{1} \; ; \; s_{2} & \textit{sequence}\\
& | \; \overline{l} = el & \textit{multiple assignment}\\
& | \; \mathbf{while} \; e \; \mathbf{do} \; s \;
| \; \mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2} & \textit{control flow}\\
& | \; \mathbf{local} \; \overline{id{:}T} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
& | \; \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
& | \; \mathbf{rec} \; id{:}T = f \; \mathbf{in} \; s & \textit{recursive function} \\
& | \; \mathbf{return} \; el & \textit{return} \\
& | \; \lfloor a \rfloor_{0} & \textit{application with no results}\\
& | \; \mathbf{fun} \; id_{1}{:}id_{2} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \textit{method declaration}\\
e ::= & & \textsc{expressions:}\\
& \;\; \mathbf{nil} & \textit{nil}\\
& | \; k & \textit{other literals}\\
& | \; id & \textit{variable access}\\
& | \; e_{1}[e_{2}] & \textit{table access}\\
& | \; {<}T{>} \; id & \textit{type coercion}\\
& | \; f & \textit{function declaration}\\
& | \; \{ \; \overline{[e_{1}] = e_{2}} \; \} \;
| \; \{ \; \overline{[e_{1}] = e_{2}},me \; \} & \textit{table constructor}\\
& | \; e_{1} + e_{2} \;
| \; e_{1} \; {..} \; e_{2} \;
| \; e_{1} == e_{2} \;
| \; e_{1} < e_{2} & \textit{binary operations}\\
& | \; e_{1} \;\&\; e_{2} \;
| \; e_{1} \; \mathbf{and} \; e_{2} \;
| \; e_{1} \; \mathbf{or} \; e_{2} & \textit{binary operations}\\
& | \; \mathbf{not} \; e \;
| \; \# \; e & \textit{unary operations} \\
& | \; \lfloor me \rfloor_{1} & \textit{expressions with one result}\\
l ::= & & \textsc{left-hand values:}\\
& \;\; id_{l} & \textit{variable assignment}\\
& | \; e_{1}[e_{2}] & \textit{table assignment}\\
& | \; id[k] \; {<}T{>} & \textit{type coercion}\\
k ::= & & \textsc{literal constants:}\\
& \;\; \mathbf{false} \; | \;
\mathbf{true} \; | \;
{\it int} \; | \;
{\it float} \; | \;
{\it string} & \\
el ::= & & \textsc{expression lists:}\\
& \;\; \overline{e} \; | \;
\overline{e}, me & \\
me ::= & & \textsc{multiple results:}\\
& \;\; a & \textit{application}\\
& | \; {...} & \textit{vararg expression}\\
a ::= & & \textsc{applications:}\\
& \;\; e(el) & \textit{function application}\\
& | \; e{:}n(el) & \textit{method application}\\
f ::= & & \textsc{function declarations:}\\
& \;\; \mathbf{fun} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \\
pl ::= & & \textsc{parameter lists:}\\
& \;\; \overline{id{:}T} \; | \;
\overline{id{:}T},{...}{:}T & \\
\end{array}
$$
\dend
\caption{The abstract syntax of Typed Lua}
\label{fig:syntax}
\end{figure*}

Figure \ref{fig:syntax} presents the abstract syntax of core Typed Lua.
It splits the syntactic categories as follows:
$s$ are statements, $e$ are expressions, $l$ are left-hand values,
$k$ are literal constants, $el$ are expression lists,
$me$ are expressions with multiple results, $a$ are function and method applications,
$f$ are function declarations, $pl$ are parameter lists,
$id$ are variable names, $T$ are first-level types, and $S$ are second-level types.
The notation $\overline{id{:}T}$ denotes the list $id_{1}{:}T_{1}, ..., id_{n}{:}T_{n}$.

Our reduced core includes two statements for declaring local variables,
one with and another without type annotations.
While we use the former to formalize how our type system handles the declaration
of annotated variables, we use the latter to formalize how our type system
handles the declaration of unannotated variables through local type inference
and also the introduction of projection types.

Our reduced core also includes a truncation operator $\lfloor \rfloor$ for
function applications, method applications, and the vararg expression.
We use $\lfloor a \rfloor_{0}$ to denote function and method applications
that produce no value, because they appear as statements.
We use $\lfloor me \rfloor_{1}$ to denote function applications,
method applications, and vararg expressions that produce only one value,
even if they return multiple values.

We also include two kinds of type coercions in our core language:
the left-hand value $id[k] \; {<}T{>}$ and the expression ${<}T{>} \;id$.
Both allow the refinement of table types.
We also split variable names into two categories to have safe aliasing
of tables in the presence of refinement.
We use $id$ when variable names appear as expressions and $id_{l}$ when
variable names appear as left-hand values.

Even though we can assign only first-level types to variables,
functions and methods can return unions of second-level types,
and our type system should be able to project these unions of
second-level types into unions of first-level types.
We use two different environments to handle this feature.
The first environment is the type environment $\env$ that maps
variables to first-level types.
We use $\env_{1}[id \mapsto T]$ to extend the environment $\env_{1}$
with the variable $id$ that maps to type $T$.
The second environment is the projection environment $\penv$ that
maps projection variables to second-level types.
We use $\penv[x \mapsto S]$ to extend the environment $\penv$
with the projection variable $x$ that maps to type $S$.
In Section \ref{sec:projections} we will show how our type system uses the
projection environment $\penv$ for handling projection types,
and also for projecting unions of second-level types into
unions of first-level types.

We present the typing rules as a deduction system for two typing relations,
one for typing statements and another for typing expressions.

We use the relation $\env_{1}, \penv \vdash s, \env_{2}$ for typing statements.
This relation means that given a type environment $\env_{1}$
and a projection environment $\penv$, we can check that a statement $s$
produces a new type environment $\env_{2}$.

We use the relation $\env_{1}, \penv \vdash e : T, \env_{2}$ for typing expressions.
This relation means that given a type environment $\env_{1}$
and a projection environment $\penv$, we can check that an expression $e$ has
type $T$ and produces a new type environment $\env_{2}$.

\subsection{Tables and refinement}
\label{sec:tables}

@@

The simplest expression involving tables in the empty table constructor.
Its type checking rule is straightforward:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR1}\\
\env_{1}, \penv \vdash \{\}:\{\}_{unique}, \env_{1}
\end{array}
\]

@@

The rule \textsc{T-CONSTRUCTOR2} uses these steps to type check a
table constructor with a non-empty list of table fields:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR2}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash ([e_{1}] = e_{2})_{i}:(K_{i},V_{i}), \env_{i+1} \\
       T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique} \;\;\;
       wf(T) \;\;\;
       n = |\;\overline{[e_{1}] = e_{2}}\;| \\
       \env_{f} = merge(\env_{1}, ..., \env_{n+1})
       \end{array}}
      {\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{f}}
\end{array}
\]

Back to our example, the table constructor
$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ has type
$\{1:``x", 2:``y", 3:``z"\}_{unique}$ through rule \textsc{T-CONSTRUCTOR2}.

As another example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
has type $\{``x":1, ``y":\{``z":2\}_{closed}\}_{unique}$ through rule
\textsc{T-CONSTRUCTOR2}.
The inner table is \emph{closed} to prevent the creation of unsafe aliases.

@@

In this example, we assign the type $\{\}_{unique}$ to the variable
$person$, then we refine its type to $\{``firstname":\String\}_{unique}$,
and then to $\{``firstname":\String, ``lastname":\String\}_{unique}$.
Rule \textsc{T-REFINE} type checks this use of refinement:
\[
\begin{array}{c}
\mylabel{T-REFINE}\\
\dfrac{\begin{array}{c}
       \env_{1}(id) = \{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n} \}_{open|unique}\\
       \env_{1}, \penv \vdash k:K, \env_{2} \;\;\;
       \not \exists i \in 1..n \; K \lesssim K_{i} \;\;\;
       V = close(T)
       \end{array}}
      {\begin{array}{l}
       \env_{1}, \penv \vdash id[k] {<}T{>}:V, \\
       \env_{2}[id \mapsto \{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}, K{:}V\}_{open|unique}]
       \end{array}}
\end{array}
\]

@@

The rules \textsc{T-IDREAD1} and \textsc{T-IDREAD2} define this non-deterministic behavior.
Rule \textsc{T-IDREAD1} uses the auxiliary function \emph{close} to
produce a \emph{closed} alias.
It also uses the auxiliary function \emph{open} to change the type of
the original reference from \emph{unique} to \emph{open},
because aliasing an \emph{unique} table type while keeping the original
reference \emph{unique} can be unsafe.
Rule \textsc{T-IDREAD2} uses the auxiliary function \emph{fix} to
produce a \emph{fixed} alias.
It also uses \emph{fix} to change the type of the original reference
to \emph{fixed}, because a \emph{fixed} table type does not allow
width subtyping.
We define these rules as follows:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-IDREAD1}\\
\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
      {\env_{1}, \penv \vdash id:close(T_{2}), \env_{1}[id \mapsto open(T_{1})]}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-IDREAD2}\\
\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
      {\env_{1}, \penv \vdash id:fix(T_{2}), \env_{1}[id \mapsto fix(T_{1})]}
\end{array}
\end{array}
\]

@@

Rule \textsc{T-IDWRITE} defines this behavior:
\[
\begin{array}{c}
\mylabel{T-IDWRITE}\\
\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = write(T_{1})}
      {\env_{1}, \penv \vdash id_{l}:close(T_{2}), \env_{1}[id \mapsto close(T_{2})]}
\end{array}
\]

@@

Our type system also has different rules for type checking table indexing to avoid
changing table types in these operations, as they cannot create aliases:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-INDEX1}\\
\dfrac{\begin{array}{c}
       \env_{1}(id) = T \;\;\;
       read(\penv, T) = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}\\
       \env_{1}, \penv \vdash e_{2}:K, \env_{2} \;\;\;
       \exists i \in 1{..}n \; K \lesssim K_{i}
       \end{array}}
      {\env_{1}, \penv \vdash id[e_{2}]:V_{i}, \env_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-INDEX2}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash e_{1}:\{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}, \env_{2}\\
       \env_{2}, \penv \vdash e_{2}:K, \env_{3} \;\;\;
       \exists i \in 1{..}n \; K \lesssim K_{i}
       \end{array}}
      {\env_{1}, \penv \vdash e_{1}[e_{2}]:V_{i}, \env_{3}}
\end{array}
\end{array}
\]

@@

Rule \textsc{T-FUNCTION1} illustrates this case:
\[
\begin{array}{c}
\mylabel{T-FUNCTION1}\\
\dfrac{\begin{array}{c}
       crall(\env_{1}[\overline{id} \mapsto \overline{T}]), \penv[\ret \mapsto S] \vdash s, \env_{2}\\
       \env_{3} = crset(\env_{1}, fav(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))\\
       \env_{4} = openset(\env_{3}, rv(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))
       \end{array}}
      {\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}T}){:}S \; s:\overline{T} \times \Void \rightarrow S, \env_{4}}
\end{array}
\]

>>>>

\begin{figure*}[!ht]
	\textbf{Abstract Syntax}\\
	\dstart
	$$
	\begin{array}{rlr}
	s ::= & & \textsc{statements:}\\
	& \;\; \mathbf{skip} & \textit{skip}\\
	& | \; s_{1} \; ; \; s_{2} & \textit{sequence}\\
	& | \; \overline{l} = el & \textit{multiple assignment}\\
	& | \; \mathbf{while} \; e \; \mathbf{do} \; s \;
	| \; \mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2} & \textit{control flow}\\
	& | \; \mathbf{local} \; \overline{id{:}T} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
	& | \; \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
	& | \; \mathbf{rec} \; id{:}T = f \; \mathbf{in} \; s & \textit{recursive function} \\
	& | \; \mathbf{return} \; el & \textit{return} \\
	& | \; \lfloor a \rfloor_{0} & \textit{application with no results}\\
	& | \; \mathbf{fun} \; id_{1}{:}id_{2} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \textit{method declaration}\\
	e ::= & & \textsc{expressions:}\\
	& \;\; \mathbf{nil} & \textit{nil}\\
	& | \; k & \textit{other literals}\\
	& | \; id & \textit{variable access}\\
	& | \; e_{1}[e_{2}] & \textit{table access}\\
	& | \; {<}T{>} \; id & \textit{type coercion}\\
	& | \; f & \textit{function declaration}\\
	& | \; \{ \; \overline{[e_{1}] = e_{2}} \; \} \;
	| \; \{ \; \overline{[e_{1}] = e_{2}},me \; \} & \textit{table constructor}\\
	& | \; e_{1} + e_{2} \;
	| \; e_{1} \; {..} \; e_{2} \;
	| \; e_{1} == e_{2} \;
	| \; e_{1} < e_{2} & \textit{binary operations}\\
	& | \; e_{1} \;\&\; e_{2} \;
	| \; e_{1} \; \mathbf{and} \; e_{2} \;
	| \; e_{1} \; \mathbf{or} \; e_{2} & \textit{binary operations}\\
	& | \; \mathbf{not} \; e \;
	| \; \# \; e & \textit{unary operations} \\
	& | \; \lfloor me \rfloor_{1} & \textit{expressions with one result}\\
	l ::= & & \textsc{left-hand values:}\\
	& \;\; id_{l} & \textit{variable assignment}\\
	& | \; e_{1}[e_{2}] & \textit{table assignment}\\
	& | \; id[k] \; {<}T{>} & \textit{type coercion}\\
	k ::= & & \textsc{literal constants:}\\
	& \;\; \mathbf{false} \; | \;
	\mathbf{true} \; | \;
	{\it int} \; | \;
	{\it float} \; | \;
	{\it string} & \\
	el ::= & & \textsc{expression lists:}\\
	& \;\; \overline{e} \; | \;
	\overline{e}, me & \\
	me ::= & & \textsc{multiple results:}\\
	& \;\; a & \textit{application}\\
	& | \; {...} & \textit{vararg expression}\\
	a ::= & & \textsc{applications:}\\
	& \;\; e(el) & \textit{function application}\\
	& | \; e{:}n(el) & \textit{method application}\\
	f ::= & & \textsc{function declarations:}\\
	& \;\; \mathbf{fun} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \\
	pl ::= & & \textsc{parameter lists:}\\
	& \;\; \overline{id{:}T} \; | \;
	\overline{id{:}T},{...}{:}T & \\
	\end{array}
	$$
	\dend
	\caption{The abstract syntax of Core Typed Lua}
	\label{fig:syntax}
\end{figure*}

Figure \ref{fig:syntax} presents the abstract syntax of 
Core Typed Lua, a large subset of concrete syntax
that the Typed Lua compiler handles. We use this abstract
syntax in the presentation of the typing rules
for tables and evolution of table types (Section~\ref{sec:tables})
and for functions with overloaded multiple returns and
projection types (Section~\ref{sec:projections}).
Core Typed Lua simplifies the presentation of typing rules
without sacrificing expressiveness.

The main differences between Core Typed Lua and the concrete
syntax are:
\begin{itemize}
\item lack of labels and goto statements, as they are incompatible
with the {\em flow typing} of Typed Lua's type system;
\item local variable declarations have an explicit block,
instead of being just statements;
\item no {\bf elseif} clause in {\bf if} statements;
\item no {\bf repeat} and {\bf for} loops, just {\bf while};
\item the only arithmetic operator is $+$, the only relational
operators are $==$ and $<$, and the only bitwise operator
is $\&$, as all other operators have typing rules identical to
one of these;
\item type coercion for evolution of table types is explicit,
with two coercion terms, the l-value $id[k] \; {<}T{>}$ for adding a field to 
a table through assignment, and the r-value ${<}T{>} \;id$ for reinterpreting
a table type to a compatible type that is expected by the context;
\item all function declarations are anonymous, so there is a
statement for declaring recursive functions;
\item function or method applications that throw away all
results or all but the first result are explicitly annotated
as such, instead this being implicit in their syntactic context;
\item identifiers that appear in l-value position are explicitly
annotated as such with an $l$ subscript.
\end{itemize}

We present the typing rules as a deduction system for two typing relations, one for typing statements and another for typing expressions. Typing judgments for statements have the form $\env_{1}, \penv \vdash s, \env_{2}$, meaning that
given a type environment $\env_{1}$ and a {\em projection environment} $\penv$ we can prove that a statement $s$ produces a new type environment $\env_{2}$. Typing judgements for
expressions have the form $\env_{1}, \penv \vdash e : T, \env_{2}$, meaning that given a type environment $\env_{1}$
and a projection environment $\penv$ we can prove that an expression $e$ has type $T$ and produces a new type environment $\env_{2}$.

The type environment is standard, and maps
identifiers to first-level types. Some typing rules of
Section~\ref{sec:tables} can change the type of an identifier,
so the typing relation has both an input and an output type
environment. The {\em projection environment} $\Pi$ maps projection labels to second-level types, and we explain
its use in Section~\ref{sec:projections}.

\subsection{Tables and Evolution of Table Types}
\label{sec:tables}

@@

The simplest expression involving tables in the empty table constructor.
Its type checking rule is straightforward:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR-EMPTY}\\
\env_{1}, \penv \vdash \{\}:\{\}_{unique}, \env_{1}
\end{array}
\]

@@

The rule \textsc{T-CONSTRUCTOR-FIELDS} uses these steps to type check a
table constructor with a non-empty list of table fields:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR-FIELDS}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash ([e_{1}] = e_{2})_{i}:(K_{i},V_{i}), \env_{i+1} \\
       T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique} \;\;\;
       wf(T) \;\;\;
       n = |\;\overline{[e_{1}] = e_{2}}\;| \\
       \env_{f} = merge(\env_{1}, ..., \env_{n+1})
       \end{array}}
      {\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{f}}
\end{array}
\]

Back to our example, the table constructor
$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ has type
$\{1:``x", 2:``y", 3:``z"\}_{unique}$ through rule \textsc{T-CONSTRUCTOR-FIELDS}.

As another example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
has type $\{``x":1, ``y":\{``z":2\}_{closed}\}_{unique}$ through rule
\textsc{T-CONSTRUCTOR-FIEDLS}.
The inner table is \emph{closed} to prevent the creation of unsafe aliases.

@@

In this example, we assign the type $\{\}_{unique}$ to the variable
$person$, then we refine its type to $\{``firstname":\String\}_{unique}$,
and then to $\{``firstname":\String, ``lastname":\String\}_{unique}$.
Rule \textsc{T-NEWFIELD} type checks this use of refinement:
\[
\begin{array}{c}
\mylabel{T-NEWFIELD}\\
\dfrac{\begin{array}{c}
       \env_{1}(id) = \{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n} \}_{open|unique}\\
       \env_{1}, \penv \vdash k:K, \env_{2} \;\;\;
       \not \exists i \in 1..n \; K \lesssim K_{i} \;\;\;
       V = close(T)
       \end{array}}
      {\begin{array}{l}
       \env_{1}, \penv \vdash id[k] {<}T{>}:V, \\
       \env_{2}[id \mapsto \{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}, K{:}V\}_{open|unique}]
       \end{array}}
\end{array}
\]

@@

The rules \textsc{T-ID-R-CLOSE} and \textsc{T-ID-R-FIX} define this non-deterministic behavior.
Rule \textsc{T-ID-R-CLOSE} uses the auxiliary function \emph{close} to
produce a \emph{closed} alias.
It also uses the auxiliary function \emph{open} to change the type of
the original reference from \emph{unique} to \emph{open},
because aliasing an \emph{unique} table type while keeping the original
reference \emph{unique} can be unsafe.
Rule \textsc{T-ID-R-FIX} uses the auxiliary function \emph{fix} to
produce a \emph{fixed} alias.
It also uses \emph{fix} to change the type of the original reference
to \emph{fixed}, because a \emph{fixed} table type does not allow
width subtyping.
We define these rules as follows:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-ID-R-CLOSE}\\
\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
      {\env_{1}, \penv \vdash id:close(T_{2}), \env_{1}[id \mapsto open(T_{1})]}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-ID-R-FIX}\\
\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
      {\env_{1}, \penv \vdash id:fix(T_{2}), \env_{1}[id \mapsto fix(T_{1})]}
\end{array}
\end{array}
\]

@@

Rule \textsc{T-ID-L} defines this behavior:
\[
\begin{array}{c}
\mylabel{T-ID-L}\\
\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = write(T_{1})}
      {\env_{1}, \penv \vdash id_{l}:close(T_{2}), \env_{1}[id \mapsto close(T_{2})]}
\end{array}
\]

@@

Our type system also has different rules for type checking table indexing to avoid
changing table types in these operations, as they cannot create aliases:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-INDEX-ID}\\
\dfrac{\begin{array}{c}
       \env_{1}(id) = T \;\;\;
       read(\penv, T) = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}\\
       \env_{1}, \penv \vdash e_{2}:K, \env_{2} \;\;\;
       \exists i \in 1{..}n \; K \lesssim K_{i}
       \end{array}}
      {\env_{1}, \penv \vdash id[e_{2}]:V_{i}, \env_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-INDEX-EXP}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash e_{1}:\{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}, \env_{2}\\
       \env_{2}, \penv \vdash e_{2}:K, \env_{3} \;\;\;
       \exists i \in 1{..}n \; K \lesssim K_{i}
       \end{array}}
      {\env_{1}, \penv \vdash e_{1}[e_{2}]:V_{i}, \env_{3}}
\end{array}
\end{array}
\]

@@

Rule \textsc{T-FUNCTION} illustrates this case:
\[
\begin{array}{c}
\mylabel{T-FUNCTION}\\
\dfrac{\begin{array}{c}
       crall(\env_{1}[\overline{id} \mapsto \overline{T}]), \penv[\ret \mapsto S] \vdash s, \env_{2}\\
       \env_{3} = crset(\env_{1}, fav(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))\\
       \env_{4} = openset(\env_{3}, rv(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))
       \end{array}}
      {\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}T}){:}S \; s:\overline{T} \times \Void \rightarrow S, \env_{4}}
\end{array}
\]
